---
title: "Latent Class Discriminant Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Class Discriminant Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The `lcda` package provides latent class discriminant analysis methods for
categorical predictors. The main functions are:

- `lcda()` for class-specific latent class models.
- `cclcda()` for common-components latent class models.
- `cclcda2()` for common-components models with class-conditional mixing weights.

All manifest variables and class labels must be integer-coded and start at 1.

## Example: CCL-CDA2 on Iris

```{r}
library(lcda)

data(iris)

iris_cat <- within(iris, {
  Sepal.Length <- as.integer(cut(Sepal.Length, breaks = c(-Inf, 5.1, 5.8, 6.4, Inf)))
  Sepal.Width <- as.integer(cut(Sepal.Width, breaks = c(-Inf, 2.8, 3.0, 3.3, Inf)))
  Petal.Length <- as.integer(cut(Petal.Length, breaks = c(-Inf, 1.6, 4.35, 5.1, Inf)))
  Petal.Width <- as.integer(cut(Petal.Width, breaks = c(-Inf, 0.3, 1.3, 1.8, Inf)))
  Species3 <- as.integer(Species)
})

model <- cclcda2(
  Species3 ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
  data = iris_cat,
  m = 1
)

model$bic
```

## Reference

Bucker, M., Szepannek, G., Weihs, C. (2010). Local Classification of Discrete
Variables by Latent Class Models. In: Locarek-Junge, H., Weihs, C. (eds)
Classification as a Tool for Research. Studies in Classification, Data Analysis,
and Knowledge Organization. Springer, Berlin, Heidelberg.
https://doi.org/10.1007/978-3-642-10745-0_13
